The imprint of global magma oceans on exoplanet demographics

### 2023-04-14 src/tex/ms.tex
brand the pattern as "habitable zone inner edge discontinuity"

% Abstract
\begin{abstract}
Key atmospheric volatiles are distributed between terrestrial planetary interiors and atmospheres during magma ocean epochs in the course of planetary accretion.
Long-term magma ocean regimes on rocky exoplanets inside the orbital runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may thus offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
Thermal stratification of super-runaway planets is expected to inflate the transit radii relative to sub-runaway planets, potentially rendering the climatic transition observationally accessible.
Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
We find that the demographic imprint of super-runaway climates should be detectable with high-precision transit photometry for planet populations $\gtrsim$ 100 planets if at least 10\% of these planets undergo a super-runaway climate.
We argue that in the near future, ESA's \plato\ mission will be the most promising survey to constrain this demographic feature.
We identify the key mission design drivers that enable a statistically sound detection, and discuss survey strategies that optimize the statistical power of combined datasets.
Observational detection of the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among low-mass exoplanets, and inform us about which stars harbor the nearest potentially habitable worlds.
\end{abstract}

@@

\subsection{Hypothesis tests}
We now turn to quantifying the ability of the simulated surveys to detect the runaway greenhouse transition and to constrain parameters of this trend.
To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
%\begin{note}
%    Bayesian model comparison requires that an alternative mode be specified against which the comparison is made.
%\end{note}
    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        H_0(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in net instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The runaway greenhouse hypothesis is then defined as
\begin{equation}\label{eq:rgh_hypo}
    H_{\mathrm{rgh}}(\theta, S) =
        \begin{cases}
            H_0, &  S \leq S_\mathrm{thresh}\\
            \langle R_\mathrm{P}\rangle (f_\mathrm{rgh},\Delta R_\mathrm{stm}, \Delta R_\mathrm{wtr}), &  S > S_\mathrm{thresh}.
        \end{cases}
\end{equation}
    Here, $f_\mathrm{rgh}$ is the fraction of planets experiencing a runaway greenhouse effect.
    $\Delta R_\mathrm{stm}$ and $\Delta R_\mathrm{wtr}$ are predicted radius changes from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

    \begin{figure*}[ht!]
%    \script{figurescript.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/plato_fgrid.pdf}
            \caption{
            Expected delta-evidences as a function of the fraction of planets with runaway greenhouse climates for different versions of the \plato\ survey.
            The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
                Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
               Even smaller samples are unlikely to yield a significant detection.
            \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the runaway greenhouse signal.
            }
            \label{fig:plato_fgrid}
        \end{centering}
    \end{figure*}
To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
For much smaller samples, only a very optimistically strong signal is likely to be detected.

@@

\subsubsection{Additional planet mass measurements}\label{sec:res_followup}
Comparing a measurement of the signal in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
With an optimistic choice of geophysical parameters (see Sect.~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.

%We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
Figure~\ref{fig:plato_fgrid} shows that investigating the habitable zone inner edge discontinuity is more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
Consequently, we achieve a statistically significant detection with smaller samples or lower dilution factors $f_\mathrm{rgh}$.
A density-based survey of 100 targets is roughly equivalent to a radius-based survey of \var{N_plato} targets.
At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred_rho}$.

\subsubsection{Dependence on host star spectral type}\label{sec:results_FGK_M}
Since the incident radiation depends on the spectral type of the host star, the relative number of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
We tested the detectability of the runaway greenhouse signal when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
%($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
The samples are volume and magnitude-limited to reflect the target counts of \plato's provisional Long-duration Observation Phase fields~\citep[$15996$ FGK stars in the P1 and P2 samples, $33948$ M~stars in the P4 sample, ][]{Nascimbeni2022}.
The resulting M~dwarf planet sample is significantly larger with $\var{N_M}\pm \var{N_M_err}$ planets compared to $\var{N_FGK}\pm \var{N_FGK_err}$ planets in the FGK sample.
No significant detection is possible in the pure FGK sample, independent of the assumed geophysical parameters.
In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.2$, similar to the case above where all spectral types are considered.

@@

Our finding that the expected trend is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Such constraints will provide insight into whether the initial water content of rocky planets is varied by systematic inter-system effects~\citep[e.g.,][]{Raymond2004,Mulders2015b,Sato2016,Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects such as planet migration~\citep{Schlecker2021} or atmospheric escape~\citep{Owen2016}.
Ultimately, such measurements will shed light on how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetary climates.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, probing the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

\subsection{Detectability of the runaway greenhouse transition}
We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
Of course, even assuming capable instrumentation and an optimal survey design, our current models may not capture fully the complexity of trends in the geophysics and demographics of small planets.
The detectability of the runaway greenhouse threshold is thus a function of both survey design and the diverse outcomes of rocky planet formation and evolution.
In the following, we will explore the key factors that may influence the emerging demographic signal.

@@

\textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water.
Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of terrestrial oceans per Gyr~\citep{2018AJ....155..195W} on even the most irradiated exoplanets~\citep[see discussion in][]{2022ApJ...938L...3L}.

@@

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the runaway greenhouse transition, since no strong dependency on instellation has been found or is expected.
A population of ``water worlds'' with low bulk densities on a wide range of orbits would merely attenuate the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals~\citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this attenuation.

@@

\subsection{Mission design trade studies}\label{sec:mission-design-trades}
To explore the impact of mission trade-offs on the detectability of the runaway greenhouse transition, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.

@@

\subsubsection{Water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the statistical radius inflation trend, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

@@

All of the above mechanisms may dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
The ``dilution factor" we introduced embodies our ignorance of these contributions.
Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.

>>>>

% Abstract
\begin{abstract}
Key atmospheric volatiles are distributed between terrestrial planetary interiors and atmospheres during magma ocean epochs in the course of planetary accretion.
Long-term magma ocean regimes on rocky exoplanets inside the orbital runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may thus offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
Thermal stratification of super-runaway planets is expected to inflate the transit radii relative to sub-runaway planets, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius/density space.
Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages contextual information from the overall planet population.
We find that the demographic imprint of super-runaway climates should be detectable with high-precision transit photometry for planet populations $\gtrsim$ 100 planets if at least 10\% of these planets undergo a super-runaway climate.
We argue that in the near future, ESA's \plato\ mission will be the most promising survey to detect and constrain habitable zone inner edge discontinuity.
We identify the key mission design drivers that enable a statistically sound detection, and discuss survey strategies that optimize the statistical power of combined datasets.
Observational detection of the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among low-mass exoplanets, and inform us about which stars harbor the nearest potentially habitable worlds.
\end{abstract}

@@

\subsection{Hypothesis tests}
We now turn to quantifying the ability of the simulated surveys to detect the habitable zone inner edge discontinuity and to constrain parameters of the runaway greenhouse transition.
To do this, we rely on a Bayesian hypothesis testing approach where we quantify the evidence of a hypothesis over another based on the (simulated) data.
For our specific problem, this implies comparing evidences for a demographic imprint of the runaway greenhouse effect to its absence.
%\begin{note}
%    Bayesian model comparison requires that an alternative mode be specified against which the comparison is made.
%\end{note}
    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        H_0(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in net instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The runaway greenhouse hypothesis is then defined as
\begin{equation}\label{eq:rgh_hypo}
    H_{\mathrm{rgh}}(\theta, S) =
        \begin{cases}
            H_0, &  S \leq S_\mathrm{thresh}\\
            \langle R_\mathrm{P}\rangle (f_\mathrm{rgh},\Delta R_\mathrm{stm}, \Delta R_\mathrm{wtr}), &  S > S_\mathrm{thresh}.
        \end{cases}
\end{equation}
    Here, $f_\mathrm{rgh}$ is the fraction of planets experiencing a runaway greenhouse effect.
    $\Delta R_\mathrm{stm}$ and $\Delta R_\mathrm{wtr}$ are predicted radius changes from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $\langle R_\mathrm{P}\rangle $.

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the habitable zone inner edge discontinuity.
Before we study limiting cases of such a detection below, we first demonstrate the interpretation process based on an optimistic scenario where the sample of characterized planets is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
Here, we assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \var{wrr} for each planet.
In this case, the habitable zone inner edge discontinuity was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

    \begin{figure*}[ht!]
%    \script{figurescript.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/plato_fgrid.pdf}
            \caption{
            Expected delta-evidences as a function of the fraction of planets with runaway greenhouse climates for different versions of the \plato\ survey.
            The median values of randomized survey simulations are shown; $\Delta \ln Z > 3$ (gray horizontal line) is considered sufficient evidence to reject the null hypothesis.
            \textit{Left:} For a large planet yield of $N = \var{N_plato}$, even small dilution factors $\sim \var{platoMinfrgh}$ allow a detection.
                A sample of 100 planets is sufficient if their masses are constrained to within \SI{10}{\percent} (dotted green line).
                Without such follow-up measurements, sufficient diagnostic power can only be achieved with this sample if $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$.
               Even smaller samples are unlikely to yield a significant detection.
            \textit{Right:} Evidences when only FGK or only M~dwarfs are considered.
                Only M~dwarfs host enough planets on both sides of the threshold instellation to allow a reliable detection of the habitable zone inner edge discontinuity.
            }
            \label{fig:plato_fgrid}
        \end{centering}
    \end{figure*}
To simulate the transit survey of ESA's \plato\ mission, we considered a volume-limited sample with a size according to projections and including all stellar spectral types.
Given \plato's expected radius precision, we find that a yield of \var{N_plato} planets enables a significant detection if the fraction of runaway greenhouse planets is larger than $ f_\mathrm{rgh} \sim \var{platoMinfrgh}$ (see Fig.~\ref{fig:plato_fgrid}).
The minimum needed fraction rises to \var{platoMinfrghHundred} for $N = 100$.
For much smaller samples, only an optimistically strong signal is likely to be detected.

@@

\subsubsection{Additional planet mass measurements}\label{sec:res_followup}
Comparing a measurement of the habitable zone inner edge discontinuity in radius space with a measurement in density space (which requires planetary mass measurements), a stronger detection occurs in the latter case:
With an optimistic choice of geophysical parameters (see Sect.~\ref{sec:res_testability}), the average measured radius change is \SI{+15}{\percent} whereas the average density change is \SI{-33}{\percent}.

%We now examine the value of a follow-up campaign to constrain planet masses with respect to the detectability of the runaway greenhouse threshold.
Figure~\ref{fig:plato_fgrid} shows that investigations of the discontinuity are more constraining when radius measurements can be augmented with mass measurements: At unchanged sample size, the difference in Bayesian evidence can be up to an order of magnitude larger.
Consequently, we achieve a statistically significant detection with smaller samples or lower dilution factors $f_\mathrm{rgh}$.
A density-based survey of 100 targets is roughly equivalent to a radius-based survey of \var{N_plato} targets.
At $N=100$, pure radius measurements require $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred}$ whereas bulk density measurements enable a detection from $f_\mathrm{rgh} \gtrsim \var{platoMinfrghHundred_rho}$.

\subsubsection{Dependence on host star spectral type}\label{sec:results_FGK_M}
Since the incident radiation depends on the spectral type of the host star, the relative number of planets on either side of $S_\mathrm{thresh}$ is different for FGK and M~dwarfs.
We tested the detectability of the habitable zone inner edge discontinuity when only FGK or only M~dwarfs are considered (see Fig.~\ref{fig:plato_fgrid}).
%($d_\mathrm{max} = \SI{75}{\parsec}, M_\mathrm{G} < 11$)
The samples are volume and magnitude-limited to reflect the target counts of \plato's provisional Long-duration Observation Phase fields~\citep[$15996$ FGK stars in the P1 and P2 samples, $33948$ M~stars in the P4 sample, ][]{Nascimbeni2022}.
The resulting M~dwarf planet sample is significantly larger with $\var{N_M}\pm \var{N_M_err}$ planets compared to $\var{N_FGK}\pm \var{N_FGK_err}$ planets in the FGK sample.
No significant detection is possible in the pure FGK sample, independent of the assumed geophysical parameters.
In the M~dwarf sample, the evidence threshold is reached around $f_\mathrm{rgh} \sim 0.2$, similar to the case above where all spectral types are considered.

@@

Our finding that the expected habitable zone inner edge discontinuity is strong except for extremely low water mass fractions gives reason for optimism regarding its detection.
It also presents a potential for constraints on the water inventory of terrestrial planets in the case of a non-detection.
Such constraints will provide insight into whether the initial water content of rocky planets is varied by systematic inter-system effects~\citep[e.g.,][]{Raymond2004,Mulders2015b,Sato2016,Lichtenberg2019,2021ApJ...913L..20L,2022ApJ...938L...3L} or if there is a predominant pattern of volatile-enrichment across planetary systems that is only modified by intra-system effects such as planet migration~\citep{Schlecker2021} or atmospheric escape~\citep{Owen2016}.
Ultimately, such measurements will shed light on how delivery and loss effects during rocky planet formation and evolution shape the diversity of exoplanetary climates.
%Should a successful detection occur, the time dependencies discussed above should be taken into account when using this population-level information to constrain exoplanet climates.
%Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
Overall, probing the runaway greenhouse discontinuity appears to be the most promising approach toward a first empirical test of the habitable zone concept.

\subsection{Detectability of the habitable zone inner edge discontinuity}
We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
Of course, even assuming capable instrumentation and an optimal survey design, our current models may not capture fully the complexity of trends in the geophysics and demographics of small planets.
The detectability of the runaway greenhouse threshold is thus a function of both survey design and the diverse outcomes of rocky planet formation and evolution.
In the following, we will explore the key factors that may influence the emerging demographic signal.

@@

\textit{Planetary evolution and duration of the steam atmosphere phase}: An inflated steam atmosphere can only be sustained until the planet has lost its water.
Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the habitable zone inner edge discontinuity.
However, the delivery uncertainty is much greater than the predicted loss rates of water by atmospheric escape, which typically is limited to in total a few tens of terrestrial oceans per Gyr~\citep{2018AJ....155..195W} on even the most irradiated exoplanets~\citep[see discussion in][]{2022ApJ...938L...3L}.

@@

\citet{Luque2022} found that small planets around orbiting red dwarfs can be classified into three density regimes with a particularly strong separation between planets consistent with a pure rocky and those consistent with a water-rich composition.
This trend does not represent a false positive scenario for the habitable zone inner edge discontinuity, since no strong dependency on instellation has been found or is expected.
A population of ``water worlds'' with low bulk densities on a wide range of orbits would merely attenuate the statistical runaway greenhouse imprint.
If the dichotomy forms primordially through migrated planets that accreted from different regions of their protoplanetary disk~\citep{Venturini2020,Burn2021,Schlecker2021,Schlecker2021b} or from inter-system variations in the desiccation of volatile-rich planetesimals~\citep{Lichtenberg2019,2021Sci...371..365L,2021ApJ...913L..20L,2022ApJ...938L...3L,2023NatAs...7...39B}, systems of all ages can be affected by this attenuation.

@@

\subsection{Mission design trade studies}\label{sec:mission-design-trades}
To explore the impact of mission trade-offs on the detectability of the habitable zone inner edge discontinuity, we simulated different survey designs and strategies and measured their capability to recover the trend and constrain its parameters.
We assessed this capability based on two determinants: the likelihood that the mission is able to detect the injected trend, and the precision with which it can constrain the parameters of that trend.

@@

\subsubsection{Water mass fraction}
The predominant mass fractions of water, which sensitively controls the atmospheric state of a rocky exoplanet, is poorly constrained.
Inferred water contents in the literature range from upper limits on the order $10^{-5}$ to ``water worlds'' with tens of percent mass fraction~\citep[e.g.,][]{Rogers2010,Unterborn2018,Mousis2020,Agol2021,Luque2022}, all of which are within the realm of theoretical predictions~\citep{Mulders2015b,Sato2016,Jin2018,Lichtenberg2019,Bitsch2019b,Venturini2020,Emsenhuber2021b,Schlecker2021,2022ApJ...938L...3L,2022ApJ...939L..19I}.
Our nominal case assumes a water mass fraction of $x_\mathrm{H_2O} = \var{wrr}$.
This can be considered a conservative choice that is unlikely to introduce a systematic overestimation of the habitable zone inner edge discontinuity, also since the hypothesis tests are only marginally sensitive on $x_\mathrm{H_2O}$ as long as it is $\gtrsim 10^{-4}$ (see Sect.~\ref{sec:res_testability}).
Cases of pure rocky composition and very low volatile contents can be considered absorbed by the dilution factor $f_\mathrm{rgh}$.
Assuming a distribution of water mass fractions instead of a fixed value would thus not significantly change our results.

@@

All of the above mechanisms may dilute the demographic imprint of the runaway greenhouse transition with contributions that are challenging to quantify.
The ``dilution factor'' we introduced embodies our ignorance of these contributions.
Importantly, while this may offset our estimates of the signal's amplitude, its shape remains the same.
